---
title: "airplane"
author: "Shahin"
date: "1/9/2020"
output: html_document
---
# 0.Introduction 

This is a study of aviation incidents and accidents. The idea ignited my mind after the recent airplane crash in Iran, Jan 2020. Before officials admit that they targetted the plane, I was speculating whether crash is frequent at the take-off stage. Then I thought why not checking the data, and see what can I learn from it? 

This is how this report is born. 

# 1. Libraries
```{r, message = F }
library(rvest)
library(tidyverse)
library(plotly)
```

# 2. Data

The data is from a related Wikipedia page, linked in the below chunk. It was a little difficult to extract it, i.e. scrape it, out of the webpage. I have never had such difficulties in scraping data from simple html pages such as wikipedia's. Anyway, as it is said, data gathering and wrangling is the most time-consuming part of any data science project. 

```{r extraction of the data}
source_url <- "https://en.wikipedia.org/wiki/List_of_aircraft_accidents_and_incidents_resulting_in_at_least_50_fatalities"

source_html <- read_html(source_url)

source_node <- 
source_html %>% 
    html_nodes(xpath = '//*[@id="mw-content-text"]/div/table[1]/tbody' )

new_colnames <- c("total_casualties","crew","passenger","nonFlying_fatalities","notes",
           "aircraft_type","incident","aircraft","location","phase","airport","distance","date")

accidents_tbl <- 
source_html %>% 
    html_nodes("#mw-content-text > div > table:nth-child(15)") %>% 
    html_table(fill = T, header = F) %>% 
    tibble::as_tibble(.name_repair = "universal") %>% 
    slice(-c(1,2))

accidents_tbl_2 <- 
    accidents_tbl %>% 
    pull(...1)

accidents_final_tbl <- 
accidents_tbl_2 %>% 
    set_names(new_colnames) %>% 
    # adding an id number to the rows 
    mutate(id = row_number()) %>% 
    
    # extract citations from the phase column
    separate(col = phase, into = c("phase","citation"), sep = "\\[" ) %>%     
    separate(col = aircraft , into = c("aircraft_type","aircraft_model"),remove = T) %>%
    mutate(phase = factor(phase)) %>%
    
    # cleaning crew, passenger  
    mutate(  crew =
               str_replace_all(crew , pattern = "\\D", "") %>%
               as.integer(),
           passenger =
               str_replace_all(passenger ,pattern = "\\D", "") %>%
               as.integer() ) %>% 
    # adding the crew and passenger to the total_casualities 
    group_by(id)  %>% 
    mutate(total_casualties = sum(crew , passenger)) %>% 
    ungroup() %>% 
    # cleaning distance 
    separate(col = distance , into = c("distance_km","rest"), sep = "\\s" , remove = T ,extra = "drop" , fill = "right" ) %>% 
    mutate(distance_km = str_replace_all(distance_km , pattern="[^\\d|\\.]", replacement = "" )) %>%
    mutate(distance_flag = str_detect(string = distance_km,pattern = "\\d")) %>% 
    mutate(distance_km = ifelse(distance_flag,distance_km,"")) %>% 
    mutate(distance_km = distance_km %>% as.numeric()) %>% 
    # feature engineering 
    mutate(date  = lubridate::as_date(date)) %>% 
    mutate(year  = lubridate::year(date) ,
           month = lubridate::month(date)) %>% 
    # making years a categorical variable of decades 
    mutate(decade = cut(year,
                        breaks = c(1930,1940,1950,1960,1970,1980,1990,2000,2010,2021),
                        labels = c("30s","40s","50s","60s","70s","80s","90s","2000-2010","2010-2020"))) %>% 
    # feature extraction from incidents 
    mutate(incident_flag = str_detect(string = incident, pattern = "\\(" )) %>% 
    mutate(airline = ifelse(incident_flag, str_split(string = incident, pattern = "\\(", simplify = T)[,1] , str_split(string = incident, pattern = "\\d", simplify = T)[,1] )) %>%
    mutate(airline = str_trim(string = airline, side = "both")) %>% 
    
    # feature extraction of location 
    mutate(location_pieces = 1+str_count(string = location, pattern = ",")) %>% 
    group_by(id) %>% 
    mutate(location_country = str_split(string = location , pattern = ",", simplify = T)[,location_pieces]) %>% 
    ungroup() %>% 
    mutate(location_country = str_replace_all(string = location_country, pattern = "[:punct:]", replacement = "")) %>% 
    mutate(location_country = str_replace_all(string = location_country, pattern = "[:digit:]", replacement = "")) %>% 
    
    # column selection 
    select(-nonFlying_fatalities, -notes, -citation, -rest, - distance_flag, -incident_flag, - location_pieces) %>% 
    select(id , everything())
    

    

#explicit_levels <- c("standing","taxi","take off","initial climb","en route","maneuvering","approach","landing","unknown")    

accidents_final_tbl %>% 
    glimpse()

accidents_final_tbl %>% 
    View()

```

So Now the data is cleaned to some extend. We can move on feature engineering. On Feature engineering, I have one principle: no compound column. Compound columns in my definition are columns with several elements of information. For instance a y+m+d date, has several elements of information, including year, month of year, season of year, day of week, day of month, day of year, day of holiday, etc. 


```{r}
    gg_barplot_incident_phase <- 
accidents_final_tbl %>% 
     filter(year != 2020) %>% 
    filter(!(is.na(phase) | is.na(decade) | phase == "" | decade == "")) %>% 
    count(phase) %>%
    mutate(total = sum(n), phase_perc = round(n / total,2)*100) %>% 
    ungroup() %>% 
    ggplot() + 
    geom_bar(aes(x = reorder(phase,phase_perc), fill = phase , y = phase_perc), stat = "identity") + 
    coord_flip() + 
    theme_light() + 
    labs( y= "Percentage of Total Incidents", x = "Flight Phases")

ggsave(gg_barplot_incident_phase , filename = "incident_phases.png", device = "png", dpi =  300)

```

```{r}
accidents_final_tbl %>% 
    filter(!(is.na(phase) | is.na(decade))) %>% 
    count(phase,decade) %>% 
    plot_ly(x = ~ phase , y = ~ decade , z = ~ n  , type = "heatmap")
```

```{r}
accidents_final_tbl %>% 
    filter(!(is.na(phase) | is.na(decade) | phase == "" | decade == "")) %>% 
    count(phase,decade) %>% group_by(decade) %>% mutate(total_decade = sum(n)) %>%
    mutate(decade_perc = round(n / total_decade,2)*100) %>% 
    arrange(decade) %>% 
    plot_ly(y = ~decade, x = ~phase, z = ~ decade_perc, type = "heatmap")


accidents_final_tbl %>% 
    filter(!(is.na(phase) | is.na(decade) | phase == "" | decade == "")) %>% 
    count(phase,decade) %>% group_by(decade) %>% mutate(total_decade = sum(n)) %>%
    mutate(decade_perc = round(n / total_decade,2)*100) %>% 
    arrange(decade) %>% 
    ggplot() + 
    geom_bar(aes(x = phase , y  = decade_perc , fill = phase) , stat = "identity") + 
    facet_grid(decade ~ .) + 
    theme(axis.text.x = element_text(color = "#993333", 
                           size = 7, angle = 45),
          axis.text.y = element_text(face = "bold", color = "blue", 
                           size = 7, angle = 45))

gg_heatmap <- 
accidents_final_tbl %>% 
    filter(year != 2020) %>% 
    filter(!(is.na(phase) | is.na(decade) | phase == "" | decade == "")) %>% 
    count(phase,decade) %>% group_by(decade) %>% mutate(total_decade = sum(n)) %>%
    mutate(decade_perc = round(n / total_decade,2)*100) %>% 
    arrange(decade) %>% 
    ggplot() + 
    geom_tile(aes(x = decade , fill  = decade_perc , y = phase)) + 
    
    theme(axis.text.x = element_text(color = "#993333", 
                           size = 7, angle = 45),
          axis.text.y = element_text(face = "bold", color = "blue", 
                           size = 7, angle = 45)) + 
    geom_text(aes(x = decade, y = phase , label = paste0(decade_perc,"%")), color = "white") + 
    theme_light()


ggsave(gg_heatmap , filename = "airplane_accidents" , device = "png" , dpi = 300)


gg_heatmap_count <- 
accidents_final_tbl %>% 
    filter(year != 2020) %>% 
    filter(!(is.na(phase) | is.na(decade) | phase == "" | decade == "")) %>% 
    count(phase,decade) %>% group_by(decade) %>% mutate(total_decade = sum(n), frequency = n ) %>%
    mutate(decade_perc = round(n / total_decade,2)*100) %>% 
    arrange(decade) %>% 
    ggplot() + 
    geom_tile(aes(x = decade , fill  = frequency , y = phase)) + 
    theme(axis.text.x = element_text(color = "#993333", 
                           size = 7, angle = 45),
          axis.text.y = element_text(face = "bold", color = "blue", 
                           size = 7, angle = 45)) + 
    geom_text(aes(x = decade, y = phase , label = frequency), color = "white") + 
    theme_light()


ggsave(gg_heatmap_count , filename = "airplane_accidents_count" , device = "png" , dpi = 300)
```

