---
title: "airplane"
author: "Shahin"
date: "1/9/2020"
output: html_document
---
# 0.Introduction 

This is a study of aviation incidents and accidents. The idea ignited my mind after the recent airplane crash in Iran, Jan 2020. Before officials admit that they targetted the plane, I was speculating whether crash is frequent at the take-off stage. Then I thought why not checking the data, and see what can I learn from it? 

This is how this report is born. 

# 1. Libraries
```{r, message = F }
library(rvest)
library(tidyverse)
library(plotly)
```

# 2. Data

The data is from a related Wikipedia page, linked in the below chunk. It was a little difficult to extract it, i.e. scrape it, out of the webpage. I have never had such difficulties in scraping data from simple html pages such as wikipedia's. Anyway, as it is said, data gathering and wrangling is the most time-consuming part of any data science project. 

```{r extraction of the data}
# the wikipedia page url 
source_url <- "https://en.wikipedia.org/wiki/List_of_aircraft_accidents_and_incidents_resulting_in_at_least_50_fatalities"

# extractaion of the table from the page: start  
source_html <- read_html(source_url)

source_node <- 
source_html %>% 
    html_nodes(xpath = '//*[@id="mw-content-text"]/div/table[1]/tbody' )



accidents_tbl <- 
source_html %>% 
    html_nodes("#mw-content-text > div > table:nth-child(15)") %>% 
    html_table(fill = T, header = F) %>% 
    tibble::as_tibble(.name_repair = "universal") %>% 
    slice(-c(1,2))

accidents_tbl_2 <- 
    accidents_tbl %>% 
    pull(...1)
# extractaion of the table from the page: end  


# new column names. They are more expressive than the original names
new_colnames <- c("total_casualties","crew","passenger","nonFlying_fatalities","notes",
           "aircraft_type","incident","aircraft","location","phase","airport","distance","date")

# data cleaning and feature engineering : start 
accidents_final_tbl <- 
accidents_tbl_2 %>% 
    set_names(new_colnames) %>% 
    # adding an id number to the rows 
    mutate(id = row_number()) %>% 
    
    # extract citations from the phase column
    separate(col = phase, into = c("phase","citation"), sep = "\\[" ) %>%     
    separate(col = aircraft ,
             into = c("aircraft_type","aircraft_model"),
             remove = T) %>%
    mutate(phase = factor(phase)) %>%
    
    # cleaning crew, passenger  
    separate(col = crew ,
             into = c("crew","crew_residuals"),
             sep = "\\D") %>%
    separate(col = passenger ,
             into = c("passenger","passenger_residuals"),
             sep = "\\D") %>%
    mutate(crew = as.integer(crew), 
           passenger = as.integer(passenger)) %>% 

    #mutate(  crew =
    #           str_replace_all(crew , pattern = "\\D", "") %>%
    #           as.integer(),
    #       passenger =
    #           str_replace_all(passenger ,pattern = "\\D", "") %>%
    #           as.integer() ) %>% 
    
    # adding the crew and passenger to the total_casualities 
    group_by(id)  %>% 
    mutate(total_casualties = sum(crew , passenger)) %>% 
    ungroup() %>% 
    # cleaning distance 
    separate(col = distance ,
             into = c("distance_km","distiance_residual"),
             sep = "\\s" ,
             remove = T ,
             extra = "drop" ,
             fill = "right" ) %>% 
    
    mutate(distance_km = str_replace_all(distance_km ,
                                         pattern="[^\\d|\\.]",
                                         replacement = "" )) %>%
    mutate(distance_flag = str_detect(string = distance_km,
                                      pattern = "\\d")) %>% 
    mutate(distance_km = ifelse(distance_flag,distance_km,"")) %>% 
    mutate(distance_km = distance_km %>% as.numeric()) %>% 
    # feature engineering on date
    mutate(date  = lubridate::as_date(date)) %>% 
    mutate(year  = lubridate::year(date) ,
           month = lubridate::month(date)) %>% 
    # making years a categorical variable of decades 
    mutate(decade = cut(year,
                        breaks = c(1930,1940,1950,1960,1970,
                                   1980,1990,2000,2010,2021),
                        labels = c("30s","40s","50s","60s","70s",
                                   "80s","90s",
                                   "2000-2010","2010-2020"))
           ) %>% 
    ## feature extraction from incidents 
    mutate(double_airline_flag =
               str_detect(string = incident,
                          pattern = " and| and |and "), 
           double_airline_flag_2 =
               str_detect(string = incident,
                          pattern = "land"), 
           double_airline_flag = ifelse(double_airline_flag_2,
                                        !double_airline_flag,
                                        double_airline_flag))%>% 
    separate(col = incident , 
             into = c("airline_1","airline_2"), 
             sep = "and" ,
             remove = F ) %>%
    mutate(airline_1 = case_when(double_airline_flag ~ airline_1,
                                 TRUE ~ incident)) %>% 
    separate(col = airline_1 , 
             into = c("airline_1","airline_3"),
             sep = ","  ) %>% 
    
    separate(col = airline_1 , 
             into = c("airline_1", "airline_1_flag"),
             sep = "\\(|\\d" ) %>% 
     separate(col = airline_2 , 
             into = c("airline_2", "airline_1_flag"),
             sep = "\\(|\\d" ) %>% 
     separate(col = airline_3 , 
             into = c("airline_3", "airline_1_flag"),
             sep = "\\(|\\d" ) %>% 

    mutate(airline_1 = str_trim(string = airline_1, side = "both"),
           airline_2 = str_trim(string = airline_2, side = "both"),
           airline_3 = str_trim(string = airline_3, side = "both")) %>% 
    
    # cleaning of some special cases that I found after eye-balling
    mutate(airline = case_when(
       str_detect(string = airline_1,
               pattern = "Aeroflot") ~ "Aeoroflot Flight",
    str_detect(string = airline_1 ,
               pattern = "Air Ferry") ~ "Air Ferry",
    str_detect(string = airline_1,
               pattern = "Algerian Air Force") ~ "Algerian Air Force",
    str_detect(string = airline_1 ,
               pattern = "Air Vietnam") ~ "Air Vietnam Flight",
    str_detect(string = airline_1 ,
               pattern = "TAC Colombia") ~ "TAC Colombia",
                            TRUE ~ airline_1
                            )) %>%
         
    # feature extraction of location
    # finding out number of pieces in each row 
    mutate(location_pieces_aux = 1+str_count(string  = location,
                                             pattern = ",")
           ) %>%
    # breaking each row to its pieces and keep the last one 
    group_by(id) %>% 
    mutate(location_country = 
               str_split(string = location ,
                        pattern = ",",
                       simplify = T)[,location_pieces_aux] %>% str_trim()
           ) %>% 
    ungroup() %>% 
    # some of the rows have 0 comma, specially when location is on the Ocean or Sea. I manually fix them 
    mutate(location_country = case_when(
        str_detect(location_country,
                   pattern = "Atlantic") ~ "Atlantic Ocean", 
        str_detect(location_country,
                   pattern = "Pacific") ~ "Pacific Ocean", 
        str_detect(location_country,
                   pattern = "Gulf of Guineaoff") ~ "Cote d'Ivoire",
        str_detect(location_country,
                   pattern = "Central America") ~ "Central America",
        str_detect(location_country,
                   pattern = "uskeeastern Mediterranean Sea") ~ "Eastern Mediterranean Sea", 
         str_detect(location_country,
                   pattern = "area of Saint Peter") ~ "France",
         str_detect(location_country,
                   pattern = "Central America") ~ "Central America",
         str_detect(location_country,
                   pattern = "DR Congo") ~ "Congo",
         str_detect(location_country,
                   pattern = "Mauritius") ~ "Mauritius",
         str_detect(location_country,
                   pattern = "Indian Ocean") ~ "Indian Ocean",
         str_detect(location_country,
                   pattern = "Namibia") ~ "Namibia",
         str_detect(location_country,
                   pattern = "Papua New Guinea") ~ "Papua New Guinea",
         str_detect(location_country,
                   pattern = "Zimbabwe") ~ "Zimbabwe",
         str_detect(location_country,
                   pattern = "Macedonia") ~ "Macedonia",
        TRUE ~ location_country)
    ) %>% 
    
    # cleaning up residuals  
    mutate(location_country = str_replace_all(string = location_country,
                                              pattern = "[:punct:]|[:digit:]",
                                              replacement = "")) %>% 
    
    # column selection 
    select(-matches(match = "_flag|_aux|_residual")) %>% 
    
    select(-nonFlying_fatalities, -notes,
           -citation,-airline, - location,
           - incident) %>% 
    
    # rearranging columns
    select(id , matches("aircraft"),
           total_casualties , 
           date:decade,
           airline_1,airline_2,airline_3,
           phase,
           location_country, everything()) %>% 
    
    # some column type correction 
    mutate(aircraft_type = aircraft_type %>% as_factor() , 
           aircract_model = aircraft_model %>% as_factor() , 
           month = month %>% as_factor() , 
           airline_1 = airline_1 %>% as_factor(), 
           airline_2 = airline_2 %>% as_factor(), 
           airline_3 = airline_3 %>% as_factor(), 
           phase = phase %>% as_factor() , 
           location_country = location_country %>% as_factor()
           )

#explicit_levels <- c("standing","taxi","take off","initial climb","en route","maneuvering","approach","landing","unknown")    

accidents_final_tbl %>% 
    glimpse()

accidents_final_tbl %>% 
    View()

```

# 3.Data Quality 

Still there may be some peculiarities lurking in the dataset. Here I check the quality. 
## 3.1 Duplications

```{r}
accidents_final_tbl %>% 
    
```

## 3.2 Extreme Values

```{r}

```

## 3.3 Illogical rows 
## 3.4 Missing Values 

# 4.EDA

Now we are at the position of doing exploratory data analysis on the dataset. Through EDA, new ideas would be emerged, that will help 

## 4.1 Uni-Variate

### 4.1.1 Casualties
```{r}
accidents_final_tbl %>% 
    ggplot() + 
    geom_histogram(aes(x = total_casualties))
```

















```{r}
    gg_barplot_incident_phase <- 
accidents_final_tbl %>% 
     filter(year != 2020) %>% 
    filter(!(is.na(phase) | is.na(decade) | phase == "" | decade == "")) %>% 
    count(phase) %>%
    mutate(total = sum(n), phase_perc = round(n / total,2)*100) %>% 
    ungroup() %>% 
    ggplot() + 
    geom_bar(aes(x = reorder(phase,phase_perc), fill = phase , y = phase_perc), stat = "identity") + 
    coord_flip() + 
    theme_light() + 
    labs( y= "Percentage of Total Incidents", x = "Flight Phases")

ggsave(gg_barplot_incident_phase , filename = "incident_phases.png", device = "png", dpi =  300)

```

```{r}
accidents_final_tbl %>% 
    filter(!(is.na(phase) | is.na(decade))) %>% 
    count(phase,decade) %>% 
    plot_ly(x = ~ phase , y = ~ decade , z = ~ n  , type = "heatmap")
```

```{r}
accidents_final_tbl %>% 
    filter(!(is.na(phase) | is.na(decade) | phase == "" | decade == "")) %>% 
    count(phase,decade) %>% group_by(decade) %>% mutate(total_decade = sum(n)) %>%
    mutate(decade_perc = round(n / total_decade,2)*100) %>% 
    arrange(decade) %>% 
    plot_ly(y = ~decade, x = ~phase, z = ~ decade_perc, type = "heatmap")


accidents_final_tbl %>% 
    filter(!(is.na(phase) | is.na(decade) | phase == "" | decade == "")) %>% 
    count(phase,decade) %>% group_by(decade) %>% mutate(total_decade = sum(n)) %>%
    mutate(decade_perc = round(n / total_decade,2)*100) %>% 
    arrange(decade) %>% 
    ggplot() + 
    geom_bar(aes(x = phase , y  = decade_perc , fill = phase) , stat = "identity") + 
    facet_grid(decade ~ .) + 
    theme(axis.text.x = element_text(color = "#993333", 
                           size = 7, angle = 45),
          axis.text.y = element_text(face = "bold", color = "blue", 
                           size = 7, angle = 45))

gg_heatmap <- 
accidents_final_tbl %>% 
    filter(year != 2020) %>% 
    filter(!(is.na(phase) | is.na(decade) | phase == "" | decade == "")) %>% 
    count(phase,decade) %>% group_by(decade) %>% mutate(total_decade = sum(n)) %>%
    mutate(decade_perc = round(n / total_decade,2)*100) %>% 
    arrange(decade) %>% 
    ggplot() + 
    geom_tile(aes(x = decade , fill  = decade_perc , y = phase)) + 
    
    theme(axis.text.x = element_text(color = "#993333", 
                           size = 7, angle = 45),
          axis.text.y = element_text(face = "bold", color = "blue", 
                           size = 7, angle = 45)) + 
    geom_text(aes(x = decade, y = phase , label = paste0(decade_perc,"%")), color = "white") + 
    theme_light()


ggsave(gg_heatmap , filename = "airplane_accidents" , device = "png" , dpi = 300)


gg_heatmap_count <- 
accidents_final_tbl %>% 
    filter(year != 2020) %>% 
    filter(!(is.na(phase) | is.na(decade) | phase == "" | decade == "")) %>% 
    count(phase,decade) %>% group_by(decade) %>% mutate(total_decade = sum(n), frequency = n ) %>%
    mutate(decade_perc = round(n / total_decade,2)*100) %>% 
    arrange(decade) %>% 
    ggplot() + 
    geom_tile(aes(x = decade , fill  = frequency , y = phase)) + 
    theme(axis.text.x = element_text(color = "#993333", 
                           size = 7, angle = 45),
          axis.text.y = element_text(face = "bold", color = "blue", 
                           size = 7, angle = 45)) + 
    geom_text(aes(x = decade, y = phase , label = frequency), color = "white") + 
    theme_light()


ggsave(gg_heatmap_count , filename = "airplane_accidents_count" , device = "png" , dpi = 300)
```

