---
title: "airplane"
author: "Shahin"
date: "1/9/2020"
output: html_document
---
# 0.Introduction 

This is a study of aviation incidents and accidents. The idea ignited my mind after the recent airplane crash in Iran, Jan 2020. Before officials admit that they targetted the plane, I was speculating whether crash is frequent at the take-off stage. Then I thought why not checking the data, and see what can I learn from it? 

This is how this report is born. 

# 1. Libraries
```{r, message = F }
library(rvest)
library(tidyverse)
library(plotly)
library(naniar)
library(UpSetR)
```

# 2. Data

The data is from a related Wikipedia page, linked in the below chunk. It was a little difficult to extract it, i.e. scrape it, out of the webpage. I have never had such difficulties in scraping data from simple html pages such as wikipedia's. Anyway, as it is said, data gathering and wrangling is the most time-consuming part of any data science project. 

```{r extraction of the data}
# the wikipedia page url 
source_url <- "https://en.wikipedia.org/wiki/List_of_aircraft_accidents_and_incidents_resulting_in_at_least_50_fatalities"

# extractaion of the table from the page: start  
source_html <- read_html(source_url)

source_node <- 
source_html %>% 
    html_nodes(xpath = '//*[@id="mw-content-text"]/div/table[1]/tbody' )



accidents_tbl <- 
source_html %>% 
    html_nodes("#mw-content-text > div > table:nth-child(15)") %>% 
    html_table(fill = T, header = F) %>% 
    tibble::as_tibble(.name_repair = "universal") %>% 
    slice(-c(1,2))

accidents_tbl_2 <- 
    accidents_tbl %>% 
    pull(...1)
# extractaion of the table from the page: end  


# new column names. They are more expressive than the original names
new_colnames <- c("total_casualties","crew","passenger","nonFlying_fatalities","notes",
           "aircraft_type","incident","aircraft","location","phase","airport","distance","date")

# data cleaning and feature engineering : start 
accidents_final_tbl <- 
accidents_tbl_2 %>% 
    set_names(new_colnames) %>% 
    # adding an id number to the rows 
    mutate(id = row_number()) %>% 
    
    # extract citations from the phase column
    separate(col = phase, into = c("phase","citation"), sep = "\\[" ) %>%     
    separate(col = aircraft ,
             into = c("aircraft_type","aircraft_model"),
             remove = T) %>%
    mutate(phase = factor(phase)) %>%
    
    # cleaning crew_casualties, passenger_casualties , nonFlying_casualties
    mutate(crew = str_replace_all(crew ,
                              pattern     = ",|est\\." ,
                              replacement = "") %>% str_trim(),
           passenger = str_replace_all(passenger ,
                                   pattern     = ",|est\\." ,
                                   replacement = "") %>% str_trim(),
           nonFlying_fatalities = str_replace_all(nonFlying_fatalities ,
                                              pattern     = ",|est\\." ,
                                              replacement = "") %>% str_trim()
           
           ) %>% 
    separate(col = crew ,
             into = c("crew_casualties","crew_residuals"),
             sep = "\\D") %>%
    separate(col = passenger ,
             into = c("passenger_casualties","passenger_residuals"),
             sep = "\\D") %>%
    separate(col = nonFlying_fatalities ,
             into = c("nonFlying_casualties","nonFlying_residuals"),
             sep = "\\D") %>%
    mutate(crew_casualties = as.integer(crew_casualties), 
           passenger_casualties = as.integer(passenger_casualties),
           nonFlying_casualties = as.integer(nonFlying_casualties)
           ) %>% 
    

    # adding the crew and passenger to the total_casualities 
    # possibly I can use rowwise()
    group_by(id)  %>% 
    mutate(total_casualties = sum(crew_casualties ,
                                  passenger_casualties,
                                  nonFlying_casualties), 
           total_flying_casualties = sum(crew_casualties ,
                                         passenger_casualties)
           ) %>% 
    ungroup() %>% 
    # cleaning distance 
    separate(col = distance ,
             into = c("distance_km","distiance_residual"),
             sep = "\\s" ,
             remove = T ,
             extra = "drop" ,
             fill = "right" ) %>% 
    
    mutate(distance_km = str_replace_all(distance_km ,
                                         pattern="[^\\d|\\.]",
                                         replacement = "" )
           ) %>%
    mutate(distance_flag = str_detect(string = distance_km,
                                      pattern = "\\d")
           ) %>% 
    mutate(distance_km = ifelse(distance_flag,distance_km,"")
           ) %>% 
    mutate(distance_km = distance_km %>% as.numeric()
           ) %>% 
    # feature engineering on date
    mutate(date  = lubridate::as_date(date)) %>% 
    mutate(year  = lubridate::year(date) ,
           month = lubridate::month(date)) %>% 
    # making years a categorical variable of decades 
    mutate(decade = cut(year,
                        breaks = c(1920,1930,1940,1950,1960,1970,
                                   1980,1990,2000,2010,2021),
                        labels = c("20s","30s","40s","50s","60s","70s",
                                   "80s","90s",
                                   "2000-2010","2010-2020"))
           ) %>% 
    ## feature extraction from incidents 
    mutate(double_airline_flag =
               str_detect(string = incident,
                          pattern = " and| and |and "), 
           double_airline_flag_2 =
               str_detect(string = incident,
                          pattern = "land"), 
           double_airline_flag = ifelse(double_airline_flag_2,
                                        !double_airline_flag,
                                        double_airline_flag))%>% 
    separate(col = incident , 
             into = c("airline_1","airline_2"), 
             sep = "and" ,
             remove = F ) %>%
    mutate(airline_1 = case_when(double_airline_flag ~ airline_1,
                                 TRUE ~ incident)) %>% 
    separate(col = airline_1 , 
             into = c("airline_1","airline_3"),
             sep = ","  ) %>% 
    
    separate(col = airline_1 , 
             into = c("airline_1", "airline_1_flag"),
             sep = "\\(|\\d" ) %>% 
     separate(col = airline_2 , 
             into = c("airline_2", "airline_1_flag"),
             sep = "\\(|\\d" ) %>% 
     separate(col = airline_3 , 
             into = c("airline_3", "airline_1_flag"),
             sep = "\\(|\\d" ) %>% 

    mutate(airline_1 = str_trim(string = airline_1, side = "both"),
           airline_2 = str_trim(string = airline_2, side = "both"),
           airline_3 = str_trim(string = airline_3, side = "both")) %>% 
    
    # cleaning of some special cases that I found after eye-balling
    mutate(airline = case_when(
       str_detect(string = airline_1,
               pattern = "Aeroflot") ~ "Aeoroflot Flight",
    str_detect(string = airline_1 ,
               pattern = "Air Ferry") ~ "Air Ferry",
    str_detect(string = airline_1,
               pattern = "Algerian Air Force") ~ "Algerian Air Force",
    str_detect(string = airline_1 ,
               pattern = "Air Vietnam") ~ "Air Vietnam Flight",
    str_detect(string = airline_1 ,
               pattern = "TAC Colombia") ~ "TAC Colombia",
                            TRUE ~ airline_1
                            )) %>%
         
    # feature extraction of location
    # finding out number of pieces in each row 
    mutate(location_pieces_aux = 1+str_count(string  = location,
                                             pattern = ",")
           ) %>%
    # breaking each row to its pieces and keep the last one 
    group_by(id) %>% 
    mutate(location_country = 
               str_split(string = location ,
                        pattern = ",",
                       simplify = T)[,location_pieces_aux] %>% str_trim()
           ) %>% 
    ungroup() %>% 
    # some of the rows have 0 comma, specially when location is on the Ocean or Sea. I manually fix them 
    mutate(location_country = case_when(
        str_detect(location_country,
                   pattern = "Atlantic") ~ "Atlantic Ocean", 
        str_detect(location_country,
                   pattern = "Pacific") ~ "Pacific Ocean", 
        str_detect(location_country,
                   pattern = "Gulf of Guineaoff") ~ "Cote d'Ivoire",
        str_detect(location_country,
                   pattern = "Central America") ~ "Central America",
        str_detect(location_country,
                   pattern = "uskeeastern Mediterranean Sea") ~ "Eastern Mediterranean Sea", 
         str_detect(location_country,
                   pattern = "area of Saint Peter") ~ "France",
         str_detect(location_country,
                   pattern = "Central America") ~ "Central America",
         str_detect(location_country,
                   pattern = "DR Congo") ~ "Congo",
         str_detect(location_country,
                   pattern = "Mauritius") ~ "Mauritius",
         str_detect(location_country,
                   pattern = "Indian Ocean") ~ "Indian Ocean",
         str_detect(location_country,
                   pattern = "Namibia") ~ "Namibia",
         str_detect(location_country,
                   pattern = "Papua New Guinea") ~ "Papua New Guinea",
         str_detect(location_country,
                   pattern = "Zimbabwe") ~ "Zimbabwe",
         str_detect(location_country,
                   pattern = "Macedonia") ~ "Macedonia",
        TRUE ~ location_country)
    ) %>% 
    
    # cleaning up residuals  
    mutate(location_country = str_replace_all(string = location_country,
                                              pattern = "[:punct:]|[:digit:]",
                                              replacement = "")) %>% 
    
    # column selection 
    select(-matches(match = "_flag|_aux|_residual")) %>% 
    
    select(-notes,
           -citation,-airline, - location,
           - incident) %>% 
    
    # rearranging columns
    select(id , matches("aircraft"),
           total_casualties , passenger_casualties,
           crew_casualties, nonFlying_casualties, 
           date:decade,
           airline_1,airline_2,airline_3,
           phase,
           location_country, everything()) %>% 
    
    # some column type correction 
    mutate(aircraft_type = aircraft_type %>% as_factor() , 
           aircract_model = aircraft_model %>% as_factor() , 
           month = month %>% as_factor() , 
           airline_1 = airline_1 %>% as_factor(), 
           airline_2 = airline_2 %>% as_factor(), 
           airline_3 = airline_3 %>% as_factor(), 
           phase = phase %>% as_factor() , 
           location_country = location_country %>% as_factor()
           )

#explicit_levels <- c("standing","taxi","take off","initial climb","en route","maneuvering","approach","landing","unknown")    

accidents_final_tbl %>% 
    glimpse()

accidents_final_tbl %>% 
    View()

```

# 3.Data Quality 

Still there may be some peculiarities lurking in the dataset. Here I check the quality. 
## 3.1 Duplications

The first step IMO is checking for duplication. Definitely, `id` was assigned based on this assumption that each row is unique, so it should not be considered for checking duplications in data. What should be considered? I don't know any exact rule. To me, it would be strange if two crashes have happened in one single day. It may happen, but in such cases, it is better to further investigate. 

```{r}
accidents_final_tbl %>% 
   count(date) %>% 
   filter(n>1)
    
```
```{r}
accidents_final_tbl %>% 
    filter(date == '1992-07-31')

# filtering out 9-11 accidents
accidents_final_tbl <- 
accidents_final_tbl %>% 
    filter(date != '2001-09-11')
```
I searched both of the incidents, and they were separate real incidents. 

The latter one is the notorious day of 9-11. I rather remove them from the dataset, since those were not unintentional accidents.

So nothing to be worried here. It is possible to go further and check duplication of other columns, but I don't see that much of value in the other columns from this aspect. For instance, the casualties, it depends on the size of the plane as usually all crew and passengers perish, and rarely any nonflying people. I tried this search and nothing unusuall was found. 


## 3.2 Extreme Values

Quantitative extreme values can be checked at column level quickly on the View() mode. 

```{r}

# for total_causalties
accidents_final_tbl %>% 
    arrange(desc(total_casualties))

accidents_final_tbl %>% 
    arrange(total_casualties)

# for distnace
accidents_final_tbl %>%
    filter(!is.na(distance_km)
           ) %>% 
    arrange(desc(distance_km))
```
It is very strange that at max, 110km the aircraft were far from the airport when the accident happened! While I expect thousands of km far from departure point accidents. 

## 3.3 Illogical rows 

This type of investigation is more difficult that mechanical evaluation of, for instance, duplicates. 
What is an illogical row here? for instance an accident with 0 crew casualties and 0 passenger casualties. 

```{r}
accidents_final_tbl %>% 
    filter(!is.na(passenger_casualties) & !is.na(crew_casualties)
           ) %>% 
    arrange(passenger_casualties,crew_casualties)
```
How is it possible? well, it is an accident in Ukraine during an aviation show, in which the bystanders were killed, but the two pilots ejected and rescued! Seems odd at first. 

Other things would be for instance accident of a USSR flight in US during cold war. Or extremely high number of passenger casualties on a small model of airplane. Or an accident date before the first fly of a specific airplane. Or a model assigned to a type of airplane which does not exists. Or when the is contraction between `phase` and `distance_km` 

## 3.4 Missing Values 

```{r}
vis_miss(accidents_final_tbl)
```

This is a very useful plot, that helps in improving the data quality. 
Some strange and some expected points are seen here. 

1. There are some missing values in `total_casualties` that seemingly happen when `crew_casualties` and `passenger_casualties` are missing but `nonFlying_casualties` are not! It may be because of the summation that I did, and forgot to use na.rm = T
2. Some records have no crew and passenger casualties, but have nonFlying_casualties. Me parece extrano! 
3. `airline_2` and `airline_3` are mostly NA values, as most of the flights had only one airline. Expectable. 
4. `distance_km` has many missing values, as expected
5. `airport` has 0% missing values! very strange to me. 

```{r}
gg_miss_var(accidents_final_tbl, show_pct = T)
gg_miss_case(accidents_final_tbl)
gg_miss_fct(x = accidents_final_tbl , fct = decade)
gg_miss_upset(accidents_final_tbl,nsets = n_var_miss(accidents_final_tbl))
```

The last plot shows that `casualties` are missing together! except for nonFlying! Also a decade is missing. let's investigate and possibly fix them. 

```{r}
accidents_final_tbl %>% 
    filter(is.na(total_casualties)
           ) %>% 
    View()
```

By checking the intact scraped dataset, I noticed that some of the records on casualties only have citations. Possibly the numbers were not agreed. 


# 4.EDA

Now we are at the position of doing exploratory data analysis on the dataset. Through EDA, new ideas would be emerged, that will help 

## 4.1 Uni-Variate

### 4.1.1 Casualties
```{r}
accidents_final_tbl %>% 
    ggplot() + 
    geom_histogram(aes(x = total_casualties))
```

















```{r}
    gg_barplot_incident_phase <- 
accidents_final_tbl %>% 
     filter(year != 2020) %>% 
    filter(!(is.na(phase) | is.na(decade) | phase == "" | decade == "")) %>% 
    count(phase) %>%
    mutate(total = sum(n), phase_perc = round(n / total,2)*100) %>% 
    ungroup() %>% 
    ggplot() + 
    geom_bar(aes(x = reorder(phase,phase_perc), fill = phase , y = phase_perc), stat = "identity") + 
    coord_flip() + 
    theme_light() + 
    labs( y= "Percentage of Total Incidents", x = "Flight Phases")

ggsave(gg_barplot_incident_phase , filename = "incident_phases.png", device = "png", dpi =  300)

```

```{r}
accidents_final_tbl %>% 
    filter(!(is.na(phase) | is.na(decade))) %>% 
    count(phase,decade) %>% 
    plot_ly(x = ~ phase , y = ~ decade , z = ~ n  , type = "heatmap")
```

```{r}
accidents_final_tbl %>% 
    filter(!(is.na(phase) | is.na(decade) | phase == "" | decade == "")) %>% 
    count(phase,decade) %>% group_by(decade) %>% mutate(total_decade = sum(n)) %>%
    mutate(decade_perc = round(n / total_decade,2)*100) %>% 
    arrange(decade) %>% 
    plot_ly(y = ~decade, x = ~phase, z = ~ decade_perc, type = "heatmap")


accidents_final_tbl %>% 
    filter(!(is.na(phase) | is.na(decade) | phase == "" | decade == "")) %>% 
    count(phase,decade) %>% group_by(decade) %>% mutate(total_decade = sum(n)) %>%
    mutate(decade_perc = round(n / total_decade,2)*100) %>% 
    arrange(decade) %>% 
    ggplot() + 
    geom_bar(aes(x = phase , y  = decade_perc , fill = phase) , stat = "identity") + 
    facet_grid(decade ~ .) + 
    theme(axis.text.x = element_text(color = "#993333", 
                           size = 7, angle = 45),
          axis.text.y = element_text(face = "bold", color = "blue", 
                           size = 7, angle = 45))

gg_heatmap <- 
accidents_final_tbl %>% 
    filter(year != 2020) %>% 
    filter(!(is.na(phase) | is.na(decade) | phase == "" | decade == "")) %>% 
    count(phase,decade) %>% group_by(decade) %>% mutate(total_decade = sum(n)) %>%
    mutate(decade_perc = round(n / total_decade,2)*100) %>% 
    arrange(decade) %>% 
    ggplot() + 
    geom_tile(aes(x = decade , fill  = decade_perc , y = phase)) + 
    
    theme(axis.text.x = element_text(color = "#993333", 
                           size = 7, angle = 45),
          axis.text.y = element_text(face = "bold", color = "blue", 
                           size = 7, angle = 45)) + 
    geom_text(aes(x = decade, y = phase , label = paste0(decade_perc,"%")), color = "white") + 
    theme_light()


ggsave(gg_heatmap , filename = "airplane_accidents" , device = "png" , dpi = 300)


gg_heatmap_count <- 
accidents_final_tbl %>% 
    filter(year != 2020) %>% 
    filter(!(is.na(phase) | is.na(decade) | phase == "" | decade == "")) %>% 
    count(phase,decade) %>% group_by(decade) %>% mutate(total_decade = sum(n), frequency = n ) %>%
    mutate(decade_perc = round(n / total_decade,2)*100) %>% 
    arrange(decade) %>% 
    ggplot() + 
    geom_tile(aes(x = decade , fill  = frequency , y = phase)) + 
    theme(axis.text.x = element_text(color = "#993333", 
                           size = 7, angle = 45),
          axis.text.y = element_text(face = "bold", color = "blue", 
                           size = 7, angle = 45)) + 
    geom_text(aes(x = decade, y = phase , label = frequency), color = "white") + 
    theme_light()


ggsave(gg_heatmap_count , filename = "airplane_accidents_count" , device = "png" , dpi = 300)
```

